---
title: "InterStressor Zooplankton - data preparation"
output: html_notebook
author: James Orr
---

Organizing the zooplankton data from the InterStressor experiment. In this notebook I'll: 

- load all the csv files 
- merge them all into one data frame
- clean up all the typos in the "class column"
- finalize the completed data that will be used for all subsequent data analyses

## Set up 

```{r, echo=TRUE, results='hide'}
# running in R version 4.3.1

#### Clear  environment 
rm(list = ls())         

#### Required packages
library(tidyverse)        # for general organisation
```

**Load data**

There are 8 folders each with ~128 csv files. For each folder I want to download all csv files, do some organizing, and then merge into one dataframe. I'll set up a function that does this and then run a loop over the 8 folders. 

```{r}
###### Function for loading data ########
data.loading <- function(path){
  csv_files <- list.files(path, pattern = ".csv", full.names = TRUE)
  temp_data_frame  <- data.frame()
  for (file in csv_files) {
    # Read the CSV file
    data <- read.csv(file, stringsAsFactors = FALSE)
    # Add a new variable with the file name
    data$filename <- basename(file)
    # Combine the data with the previously read files
    temp_data_frame <- bind_rows(temp_data_frame, data)
  }
  return(temp_data_frame)
  }

# Run the function for all eight folders
x2_T1_data <- data.loading("data/input/x2/T1_13_05_22/")
x2_T2_data <- data.loading("data/input/x2/T2_18_05_22/")
x2_T4_data <- data.loading("data/input/x2/T4_27_05_22/")
x2_T5_data <- data.loading("data/input/x2/T5_01_06_22/")

x4_T1_data <- data.loading("data/input/x4/T1_13_05_22/")
x4_T2_data <- data.loading("data/input/x4/T2_18_05_22/")
x4_T4_data <- data.loading("data/input/x4/T4_27_05_22/")
x4_T5_data <- data.loading("data/input/x4/T5_01_06_22/")
```

**General cleaning**

Now for each of these dataframes I'll add in a date, I'll remove Particle ID and I'll extract the mesocosm ID from the file name (it'll be the sX before the first underscore).

```{r}
###### Function for general cleaning ########
data.cleaning <- function(data, date, lens){
  temp_data <- data %>%
    # create date variable
    mutate(date = date) %>%
    # create a lens variable
    mutate(lens = lens) %>%
    # get rid of Particle.ID
    select(-Particle.ID) %>%
    # extract everything before the first "_" in "filename" (this is mesocosm ID)
    mutate(meso = as.numeric(str_remove(str_extract(.$filename, "^[^_]+"), "^s")))
  return(temp_data)
}

# Run the function for all eight data frames
x2_T1_data <- data.cleaning(x2_T1_data, "13/05/2022", "x2")
x2_T2_data <- data.cleaning(x2_T2_data, "18/05/2022", "x2")
x2_T4_data <- data.cleaning(x2_T4_data, "27/05/2022", "x2")
x2_T5_data <- data.cleaning(x2_T5_data, "01/06/2022", "x2")

x4_T1_data <- data.cleaning(x4_T1_data, "13/05/2022", "x4")
x4_T2_data <- data.cleaning(x4_T2_data, "18/05/2022", "x4")
x4_T4_data <- data.cleaning(x4_T4_data, "27/05/2022", "x4")
x4_T5_data <- data.cleaning(x4_T5_data, "01/06/2022", "x4")
```

**Combine data frames**

```{r}
all_data <- rbind(x2_T1_data, x2_T2_data, x2_T4_data, x2_T5_data,
                  x4_T1_data, x4_T2_data, x4_T4_data, x4_T5_data)
rm(x2_T1_data, x2_T2_data, x2_T4_data, x2_T5_data,
   x4_T1_data, x4_T2_data, x4_T4_data, x4_T5_data)
```

**Join in treatment data**

```{r}
interstressor_design <- read.csv("data/input/interstressor-design.csv")
all_data <- left_join(all_data, interstressor_design, by = "meso")
rm(interstressor_design)
```

**Clean up the ID names** 

There are quite a few duplicates/typos in the class column: 

```{r}
table(all_data$Class)
```
I'll go through this table and clean it up 

```{r}
# first set everything to lower case
all_data$Class <- tolower(all_data$Class)

# go through taxa and standardize
all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_alona", 
                                                               "4x_alona"), 
                         "alona")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_alona_abun", 
                                                               "2x_alona_abundance", 
                                                               "4x_alona_abundance"), 
                         "alona_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_chydorus", 
                                                               "4x_chydorus"), 
                         "chydorus")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_chydorus_abun", 
                                                               "2x_chydorus_abundance",
                                                               "4x_chydorus_abun",
                                                               "4x_chydorus_abundance"), 
                         "chydorus_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_copepod_cyclopoids", 
                                                               "2x_copepods_cyclopoid",
                                                               "2x_copepods_cyclopoids",
                                                               "4x_copepods_cyclopoids"), 
                         "cyclopoida")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_copepod_cyclopoids_abun", 
                                                               "2x_copepods_ciclopoids_abundance",
                                                               "2x_copepods_cyclopoid_abun",
                                                               "4x_copepods_cyclopoids_abundance"), 
                         "cyclopoida_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_copepods_calanoids", 
                                                               "2x_copepods_calenoid"), 
                         "calanoida")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_copepods_calenoid_abun"), 
                         "calanoida_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_daphnia",
                                                               "2x_daphnia_magna",
                                                               "2x_daphnia_sp"), 
                         "daphnia")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_daphnia_abun",
                                                               "2x_daphnia_abundance"), 
                         "daphnia_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_ephippia",
                                                               "4x_ephippia"), 
                         "ephippia")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_ephippia_abun",
                                                               "2x_ephippia_abundance",
                                                               "4x_ephippia_abun"), 
                         "ephippia_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_hexarthra",
                                                               "2x_hexarthra_abun",
                                                               "2x_hexarthra_abundance",
                                                               "2x_hexartra_abundance",
                                                               "unknown"), 
                         "unknown/hexarthra")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_keratella_cochlearis",
                                                               "4x_keratella_c"), 
                         "keratella_cochlearis")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("4x_keratella_c_abun",
                                                               "4x_keratella_cochlearis_abun"), 
                         "keratella_cochlearis_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_keratella_quadrata",
                                                               "4x_keratella_q"), 
                         "keratella_quadrata")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_keretella_quadrata_abun",
                                                               "4x_k_quadrata_abundance",
                                                               "k_quadrata_abundance"), 
                         "keratella_quadrata_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_naupli",
                                                               "4x_nauplii"), 
                         "nauplius")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_nauplii_abun",
                                                               "2x_nauplii_abundance",
                                                               "2x_nauplii_top_side",
                                                               "4x_nauplii_abundance"), 
                         "nauplius_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_ostracods",
                                                               "4x_ostracods"), 
                         "ostracoda")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_ostracods_abundance",
                                                               "ostracods_abundance"), 
                         "ostracoda_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_pleuroxus"), 
                         "pleuroxus")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_scapholeberis",
                                                               "4x_scapholeberis"), 
                         "scapholeberis")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_scapholeberis_abun",
                                                               "2x_scapholeberis_abundance",
                                                               "4x_scapholeberis_abundance"), 
                         "scapholeberis_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_simocephalus"), 
                         "simocephalus")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("2x_simocephalus_abun",
                                                               "2x_simocephalus_abundance"), 
                         "simocephalus_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("4x_brachionus"), 
                         "brachionus")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("4x_cydoridae"), 
                         "chydoridae")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("4x_lecane"), 
                         "lecane")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("4x_lepadella"), 
                         "lepadella")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("4x_lepadella_abundance"), 
                         "lepadella_abun")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("4x_polyarthra"), 
                         "polyarthra")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("4x_testudinella"), 
                         "testudinella")

all_data$Class <-replace(all_data$Class, all_data$Class %in% c("4x_vorticella"), 
                         "vorticella")
```

Should be all good now. Classes with the "_abun" mean that they can only be used for abundance data (i.e., not biomass)

```{r}
table(all_data$Class)
```
**Create a "Taxa" variable**

Combine the classes of the same taxon (with and without "_abun") - to obtain the abundance of each taxon 

```{r}
all_data$Taxa <- all_data$Class 

# Use gsub to remove "_abun" from the column
all_data$Taxa <- gsub("_abun", "", all_data$Taxa)


# option to remove polyarthra and unkown/hexarthra (this is aquatic vascular plant hair)
all_data <- all_data %>%
  filter(Taxa != "polyarthra") %>%
  filter(Taxa != "unknown/hexarthra")

table(all_data$Taxa)

```

**Add in Microscope data**

These were the zooplankton (n = 101) that were too large to process in the flowcam. 

```{r}
# load the data
micro <- read.csv("data/input/microscope/zooplankton_microscope>1mm.csv", header = T)

# prepare the dataset 

micro <- micro %>%
  # select the columns we want
  select(c(Date, Mesocosm, Species, 
           C.Length..um., C.Width..um.)) %>%
  # select the rows we want 
  filter(Species != "N/A") %>%
  filter(Species != "Empty")

# fix species names 
micro$Species <-replace(micro$Species, micro$Species %in% c("Cyclopoid"), 
                         "cyclopoida")
micro$Species <-replace(micro$Species, micro$Species %in% c("Daphnia"), 
                         "daphnia")
micro$Species <-replace(micro$Species, micro$Species %in% c("Ephippia\xca"), 
                         "ephippia")
micro$Species <-replace(micro$Species, micro$Species %in% c("Scap"), 
                         "scapholeberis")
micro$Species <-replace(micro$Species, micro$Species %in% c("Simo"), 
                         "simocephalus")

# standardize column names 
micro <- micro %>%
  rename(date = Date,
         meso = Mesocosm,
         Class = Species,
         Length = C.Length..um.,
         Width = C.Width..um.) %>%
  mutate(Taxa = Class) %>%
  mutate(filename = "zooplankton_microscope>1mm.csv") %>%
  mutate(lens = "microscope")

# add in experimental design data
interstressor_design <- read.csv("data/input/interstressor-design.csv")
micro <- left_join(micro, interstressor_design, by = "meso")
rm(interstressor_design)

# finally join this micro to the all_data 
all_data <- rbind(all_data, micro)
rm(micro)
```


**Save the full dataset**

```{r}
write.csv(all_data, "data/output/interstressor_zoo_raw.csv", row.names = F)
```


